---
title: "BCB420_Assignment3"
subtitle: "Data set Pathway and Network Analysis"
author: "Mingzheng liu"
date: "04/04/2022"
output:
  html_document:
    toc: yes
    toc_depth: '2'
    df_print: paged
  html_notebook:
    toc: yes
    toc_depth: 2
  pdf_document:
    fig_caption: yes
bibliography: Assignment.bib
always_allow_html: yes
---
## Objective of Assignment 3
* To take the ranked data that was created in Assignment #2 
* To perform non-thresholded pathway analysis
* To summarize and visualize your results using Cytoscape and the Enrichment map pipeline.

# Introduction
## General Background
The dataset is derived from genetically engineered Foxd1Cre;Smo(flox/-) house mice and aims to investigate the mRNA content of mutant kidneys to gain a deeper understanding of the molecular mechanisms underlying the developmental defects observed in the kidney. The Hedgehog-GLI pathway, which is responsible for the development and patterning of various tissues and organs, including the kidney, interacts with the transforming growth factor beta (TGF) signaling pathway, another critical regulator of kidney development. Smo, a key component of the Hedgehog-GLI signaling pathway, plays a crucial role in human kidney development and modulates TGF&beta; signaling in the kidney, and its disruption can lead to abnormal kidney development and function. The ultimate goal of understanding the role of Smo in these signaling pathways is to uncover the molecular mechanisms underlying normal kidney development and diseases, which could ultimately lead to the development of new therapeutic strategies for kidney disorders. (@Rowan2018)

## Previous Work
In the previous work on the dataset with accession ID GSE103923, an initial processing of the data has been done, including accessing the dataset quality, filtering the low expression gene. The dataset was downloaded in a normalized data with identified as a mixed of HUGO symbol and he Mouse Genome Informatics (MGI) symbol. Hence, a identifier mapping was performed, and no extra normalization was applied. The result of assignment one is a clean, normalized dataset stored in the root directory as "smo_exp_normalized.txt". The downloaded cleaned annotated dataset with calculated fold change and adjusted p-value is stored in the root directory as "smo_exp_annot.txt"
* Assignment 1 in htmlViewer can be found [here](https://htmlpreview.github.io/?https://github.com/bcb420-2023/Mingzheng_Liu/blob/main/Assignment1.html)
In the secound stage, a differential expression analysis is performed on the dataset. The genes were ranked according to differential expression. A thresholded over-representation analysis is performed to highlight dominant themes in the top set of genes. Three sources of annotastions, Gene Ontology Biological Process (GO:BP), Reactome (REAC), and WikiPathways (WP) from g:profile are used for this analysis. The organism is specified as “mmusculus”, since the analysis is being performed on mouse genes for simulating human physiology process. The version of the annotation database is the most current version available at the time of running the analysis, as the gost function retrieves annotation data from the g:Profiler web service, which is regularly updated. For all genesets, 107 genesets were return, among which 69 genesets are up-regulated and 31 genesets are down-regulated.
* Assignment 2 in htmlViewer can be found [here](https://htmlpreview.github.io/?https://github.com/bcb420-2023/Mingzheng_Liu/blob/main/A2_MingzhengLiu.html)
 
## Preparation
Tools are needed for the GSEA analysis: 
* Packages used
* R (@r)
* [BiocManager](https://CRAN.R-project.org/package=BiocManager) (@morgan2021a)
* [knitr](https://yihui.org/knitr/) (@knitr)
* [ggplot2](https://ggplot2.tidyverse.org/) (@ggplot2)
* [RCurl](https://cran.r-project.org/web/packages/RCurl/index.html) (@rcurl)
* [RCy3](https://bioconductor.org/packages/release/bioc/html/RCy3.html) (@RCy3)

```{r Repare packages, message=FALSE, warning=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")
library(BiocManager)
if (!requireNamespace("knitr", quietly = TRUE))
  install.packages("knitr")
library(knitr)
if (!requireNamespace("ggplot2", quietly = TRUE))
  install.packages("ggplot2")
library(ggplot2)
if (!requireNamespace("RCurl", quietly = TRUE)){
  install.packages("RCurl")}
library(RCurl)
if (!requireNamespace("RCy3", quietly = TRUE)){
  BiocManager::install("RCy3")}
library(RCy3)
```

## Prepare dataset from Assignment 2
###Step 1: Prepare gmt file
```{r gmt Prepare}
# Download mouse GO to retrieve the gmt file
# Since we are using a mouse model, we use mouse GO
# get the current working directory path
data_dir <- paste0(getwd(), "/")
# Set url of baderlab to retrieve mouse GO
gmt_url = "http://download.baderlab.org/EM_Genesets/current_release/Mouse/symbol/Mouse_GO_AllPathways_no_GO_iea_March_02_2023_symbol.gmt"
# retrieve mouse GO
if (!file.exists("Mouse_GO.gmt")) {
  download.file(gmt_url, "Mouse_GO.gmt")
}
# get the gmt file path
dest_gmt_file <- paste0(data_dir, "Mouse_GO.gmt")
```
###Step 2: Prepare rnk file
```{r rnk Prepare}
# Since the dataset is pre-normalized with logFC and adjusted p-value
# We make the rnk file from the cleaned dataset

# Load the differential expression analysis results from previous work
de_results <- read.table("smo_exp_annot.txt", header=TRUE)

# calculate rank for each gene
de_results[,"rank"] <- -log(de_results$padj, 10) * sign(de_results$log2FoldChange) 
de_results <- de_results[order(de_results$rank),]

# Combine the rank, gene IDs
rnk <- cbind(de_results$id, de_results$rank)
colnames(rnk) <- c("GeneID", "rank")

# Save the rnk file
if (!file.exists("smo_exp.rnk")) {
write.table(rnk, file="smo_exp.rnk", quote=FALSE, sep="\t", row.names=FALSE)
}
```

# Non-thresholded Gene set Enrichment Analysis (GSEA)
## Configurations for running the GSEA
```{r GSEA configuring}
# Set the path to the GSEA jar file
# The GSEA jar file is required to run GSEA automatically. 
# the gsea jar file is uploaded to the main directory.

# In order to run GSEA automatically, I need to specify the path to the gsea jar file.
# gsea jar file is uploaded to the main directory
# since in the latest release of gsea (4.0.2) they no longer release a bundled jar
gsea_jar <- paste0(data_dir,"GSEA_4.3.2/gsea-cli.sh")

# set the flag to run gsea later
run_gsea = FALSE # set back from TRUE to FALSE once I have run gsea

# set the working directory
working_dir <- getwd()

# set the gsea blank for the notebook to discover the gsea directory for itself
gsea_directory = ""
# set analysis name
analysis_name <- "Smo_KO_vs_WT"

# The ranked file from assignment 2 is uploaded in the current directory named "smo_exp.rnk"
rnk_file <- "smo_exp.rnk"
expression_file <- "smo_RNASeq_expression.gct"
classes_file <- "smo_RNASeq_classes.cls"
```

## Run GSEA
```{r run GSEA, message=FALSE, warning=FALSE}
# since we using gsea 4.3.2
# we run the gsea from r here
if(run_gsea){
  command <- paste(
    "",gsea_jar,  
    "GSEAPreRanked -gmx", dest_gmt_file, 
    "-rnk" ,file.path(working_dir,rnk_file), 
    "-collapse false -nperm 1000 -scoring_scheme weighted -rpt_label ",analysis_name,
    "  -plot_top_x 20 -rnd_seed 12345  -set_max 200 -set_min 15 -zip_report false -out" ,working_dir, 
    " > gsea_output.txt",sep=" ")
  system(command)
}
```

## Get the name of the GSEA output directory
```{r GSEA output naming}
# find the gsea_directory
if(gsea_directory == ""){
  # find the directories of gsea result
  gsea_directories <- list.files(path = working_dir, pattern = "\\.GseaPreranked")
  # get the details on the files
  details = file.info(file.path(getwd(),gsea_directories))
  # order according to newest to oldest
  details = details[with(details, order(as.POSIXct(mtime),decreasing = TRUE)), ]
  # use the newest file
  gsea_output_dir <- row.names(details)[1]
} 
```

# Visualize Gene set Enrichment Analysis in Cytoscape
## Launch Cytoscape
```{r Cytoscape launching}
# use functions from RCy3 package
# use with cytoscape app open otherwise get error
# check accept connection when open cytoscape
cytoscapePing() 
cytoscapeVersionInfo() # cytoscape info
```

## Create an Enrichment map
```{r EM set-up}
# Set up all parameters in the em command
# path to the baderlab gmt dataset used for GSEA
gmt_gsea_file <-  dest_gmt_file
# defined threshold for GSEA enrichments (need to be strings for cyrest call)
pvalue_gsea_threshold <- "1"
qvalue_gsea_threshold <- "0.05"
similarity_threshold <- "0.375"
similarity_metric = "COMBINED" 
cur_model_name <- analysis_name
# path to GSEA result
gsea_results_path <- file.path(gsea_output_dir,"edb")
gsea_results_filename <- file.path(gsea_results_path,"results.edb")
# path to the rank file
gsea_ranks_file <- file.path(gsea_results_path,list.files(gsea_results_path,pattern=".rnk"))
```

```{r EM Creation}
#create EM
current_network_name <- paste(cur_model_name,pvalue_gsea_threshold,qvalue_gsea_threshold,sep="_")
# EM command
em_command = paste('enrichmentmap build analysisType="gsea" gmtFile=', gmt_gsea_file,
                   'pvalue=',pvalue_gsea_threshold, 'qvalue=',qvalue_gsea_threshold,
                   'similaritycutoff=',similarity_threshold,
                   'coefficients=',similarity_metric,'ranksDataset1=', 
                   gsea_ranks_file,'enrichmentsDataset1=',gsea_results_filename, 
                   'filterByExpressions=false')
# get the result of suid or "Failed" status
response <- commandsGET(em_command)
current_network_suid <- 0
# check response 
if(grepl(pattern="Failed", response)){
  paste(response)
} else {
  current_network_suid <- response
}
```

```{r network naming}
#check to see if the network name is unique
current_names <- getNetworkList()
if(current_network_name %in% current_names){
  #if the name already exists in the network names then put the SUID in front
  # of the name (this does not work if you put the suid at the end of the name)
  current_network_name <- paste(current_network_suid,current_network_name, sep="_")
}
```

## Get a screen shot of the initial network.
```{r EM screenshot}
fitContent()
output_network_file <- file.path(getwd(), "initial_screenshot_network.png")
if(file.exists(output_network_file)){
  # cytoscape hangs waiting for user response if file already exists.
  # Remove it first
  response <- file.remove(output_network_file)
} 
# put the new screenshot
response <- exportImage(output_network_file, type = "png")
```

```{r image display, out.width = "90%", echo = FALSE, message =FALSE, fig.cap="Figure 1. Initial Enrichment Map of differentially expressed genes in E14.5 Foxd1Cre;Smo(flox/-) house mice kidney cells. The gene sets were derived from the Molecular Signatures Database (MSigDB), and the analysis was performed using GSEA software. Each node represents a gene set. The color indicates the enrichment score, where red represents enrichment and blue represents depletion. Gene sets with similar functions are grouped together, and the size of each circle represents the number of genes in the set. The map shows the significant enrichment of gene sets related to cell mitosis, highlighting their role in kidney development."}
htmltools::img(src = knitr::image_uri(output_network_file), 
               alt = 'Initial Enrichment Map', 
               style = 'margin:0px auto;display:block')
```
# Interpretation and detailed view of results
1. Do the enrichment results support conclusions or mechanism discussed in the original paper? How do these results differ from the results you got from Assignment #2 thresholded methods

2. Can you find evidence, i.e. publications, to support some of the results that you see. How does this evidence support your result?

3. Choose a specific pathway or theme to investigate in more detail. Why did you choose this pathway or theme? Show the pathway or theme as a gene network or as a pathway diagram. Annotate the network or pathway with your original log fold expression values and p-values to show how it is effected in your model. (Hint: if the theme or pathway is not from database that has detailed mechanistic information like Reactome you can use apps like GeneMANIA or String to build the the interaction network.)

# References