---
title: "BCB420_Assignment"
author: "Mingzheng liu"
date: "30/01/2022"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---
 BCB420 Assignment 1 Data Selection and Initial Processing
---
## Overview
* where to find expression data? - Gene Expression Omnibus (GEO)
* tools for mining data - GEOmetadb package
* different platforms available
* how to find a dataset of interest?

### Preparation
We need some tools for downloading the datase from GEO: The GEOmetadb package
We download GEOmetadb through BioManager.

```{r}
# if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
# if (!requireNamespace("GEOmetadb", quietly = TRUE)) BiocManager::install("GEOmetadb")
library(BiocManager)
library(GEOmetadb)
```
GEOmethdb Set-up
* get the meta data
* package download the meta data not the expression data
```{r}
# Check if the database name GEOmetadb is available
if(!file.exists('GEOmetadb.sqlite')) getSQLiteFile()
# The database version
file.info('GEOmetadb.sqlite')

# connect GEO meta data database by absolute path
con <- dbConnect(SQLite(),'GEOmetadb.sqlite')

```

### 1. Select an Expression Data Set
* find the dataset of interests
```{r}
gse <- GEOquery::getGEO("GSE103923",GSEMatrix=FALSE)
knitr::kable(data.frame(head(GEOquery::Meta(gse))), format = "html")

current_gpl <- names(GPLList(gse))[1]
current_gpl_info <- Meta(getGEO(current_gpl))
current_gpl_info$title
current_gpl_info$last_update_date
current_gpl_info$organism
```

### 2. Clean the data and map to HUGO symbols
```{r}
```

### 3. Apply Normalization
```{r}
```

### 4. Interpret, and document
```{r}
```


