---
title: "BCB420_Assignment"
author: "Mingzheng liu"
date: "30/01/2022"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---
 BCB420 Assignment 1 Data Selection and Initial Processing
---
## Overview
* where to find expression data? - Gene Expression Omnibus (GEO)
* tools for mining data - GEOmetadb package
* different platforms available
* how to find a dataset of interest?

### Preparation
We need some tools for downloading the datase from GEO: The GEOmetadb package
We dowload GEOmetadb through BioManager.

```{r}
# if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
# if (!requireNamespace("GEOmetadb", quietly = TRUE)) BiocManager::install("GEOmetadb")
# if (!requireNamespace("ggplot2", quietly = TRUE)) install.packages("ggplot2")
library(BiocManager)
library(GEOmetadb)
library(ggplot2)
```
* GEOmethdb Set-up
** get the meta data
** package download the meta data not the expression data
```{r}
if(!file.exists('GEOmetadb.sqlite')) getSQLiteFile()
file.info('GEOmetadb.sqlite')

# connect Geo meta data db
con <- dbConnect(SQLite(),'GEOmetadb.sqlite')

# check out the tables available
geo_tables <- dbListTables(con)
geo_tables

# details of an individual table
dbListFields(con,'gse')

# Run SQL queries
results <- dbGetQuery(con,'select * from gpl limit 5')
knitr::kable(head(results[,1:5]), format = "html")
```
* Platforms check
```{r}
# check out the platforms
num_platforms <- dbGetQuery(con,'select count(*) from gpl')
num_platforms

# other informations we can get from another platform GPL table
dbListFields(con,'gpl')


```

```{r}
```

### 1. Select an Expression Data Set
```{r}
```

### 2. Clean the data and map to HUGO symbols
```{r}
```

### 3. Apply Normalization
```{r}
```

### 4. Interpret, and document
```{r}
```


