---
title: "BCB420_Assignment1"
author: "Mingzheng liu"
date: "14/02/2022"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---
 BCB420 Assignment 1 Data Selection and Initial Processing
---
# Overview
* where to find expression data? - Gene Expression Omnibus (GEO)
* tools for mining data - GEOmetadb package
* different platforms available
* how to find a dataset of interest?

# Preparation
We need some tools for downloading the datase from GEO: The GEOmetadb package
We download GEOmetadb through BioManager.

```{r}
if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
library(BiocManager)
if (!requireNamespace("GEOmetadb", quietly = TRUE)) BiocManager::install("GEOmetadb")
library(GEOmetadb)
# we need knitr to present the data
library(knitr)
```
###GEOmethdb Set-up
* get the meta data
* package download the meta data not the expression data
```{r}
# Check if the database name GEOmetadb is available
if(!file.exists('GEOmetadb.sqlite')) getSQLiteFile()
# The database version
file.info('GEOmetadb.sqlite')

# connect GEO meta data database by absolute path
con <- dbConnect(SQLite(),'GEOmetadb.sqlite')
```

# 1. Select an Expression Data Set
* find the dataset of interests
* Particularly, I am interested in the kidney embryology.
```{r}
gse <- getGEO("GSE103923",GSEMatrix=FALSE)
kable(data.frame(head(GEOquery::Meta(gse))), format = "html")
```
* information about Platform
```{r}
current_gpl <- names(GPLList(gse))[1]
current_gpl_info <- Meta(getGEO(current_gpl))
current_gpl_info$title
current_gpl_info$last_update_date
current_gpl_info$organism
```
* Get the expression Data
```{r}
sfiles = getGEOSuppFiles('GSE103923')
fnames = rownames(sfiles)

smo_exp = read.delim(fnames[1],header=TRUE,check.names = FALSE)

read.table(fnames[1],header=TRUE,check.names=FALSE,nrows = 1)[1,1:4]

read.table(fnames[1],header=TRUE,check.names=TRUE,nrows = 1)[1,1:4]

kable(ca125_exp[1:15,1:5], format = "html")

```

# 2. Clean the data and map to HUGO symbols


```{r}
```

# 3. Apply Normalization
```{r}
```

# 4. Interpret, and document
```{r}
```

# Reference

