---
title: "BCB420_week7"
subtitle: "Subtitle goes here"
author: "Mingzheng liu"
date: "28/02/2022"
output: 
  html_document:
    toc: true
    toc_depth: 2
bibliographt: lec7.bib
csl: biomed-central.csl    
editor_options: 
  markdown: 
    wrap: 72
    
---
 BCB420 Lecture 7 Notes - Rmarkdown Tips
---
# Overview
* Table of contents (TOC)
* Code chunk specifications
* References using bibtex

# Extra Reading:
* https://www.rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf
* https://yihui.org/knitr/

# TOC
TOC will be generated by taking all entries specified as a heading using "#"
Depending on what depth you specify this will only include the number of "#" specified

# Code Chunk specif.
* message = FALSE - to get rid of all those info messages that make the report un-readable. This option only suppresses messages; it does not affect the output or errors generated by the code chunk.
```{r my-code-chunk1, message = FALSE}
# This code chunk generates some output but suppresses messages
x <- 5
x
library(biomaRt)
installed.packages(biomartCacheInfo())
```

* warning = FALSE - Some warning may indicate that a function is not being used correctly. 
```{r my-code-chunk2, warning = FALSE}
# This code chunk generates some warnings but suppresses them

# Define a function that may generate a warning
my_function <- function(x) {
  if (x < 0) {
    warning("x is negative!")
  }
  return(x^2)
}

# Call the function with a negative argument
result <- my_function(-5)

# Display the result
print(result)
```

In this example, the warning = FALSE option is included in the code chunk header. This option tells R Markdown to suppress any warnings that are generated by the code chunk. In this case, the my_function() function generates a warning when called with a negative argument, but that warning is not displayed in the output.

* echo and eval are both useful for reports where you don't want to show the code you are using. If you are giving this report to a biologist or experimentalist they don't want to see the code that generated the data or figures.
```{r my-code-chunk3, echo = TRUE, eval = FALSE}
# This code chunk includes the code but does not evaluate it

# Define a function
my_function <- function(x) {
  return(x^2)
}

# Display the function
my_function
```
In this example, the echo = TRUE option is included in the code chunk header, which means that the code in the chunk will be displayed in the output. However, the eval = FALSE option is also included, which means that the code will not actually be evaluated when the document is rendered. Instead, the code will be displayed as plain text in the output.
Note that if you set echo = FALSE and eval = FALSE, the code chunk will not be displayed in the output at all:
```{r my-code-chunk4, echo = FALSE, eval = FALSE}
# This code chunk is not displayed in the output

# Define a function
my_function <- function(x) {
  return(x^2)
}

# Display the function
my_function
```
In this example, the code chunk will not be displayed in the output because both echo = FALSE and eval = FALSE are set.

# Reference using bibtext
* Step 1: Create a separate file with the .bib extension that contains the bibliographic information for your references.
example :
```{}
@article{cytoscape2003,
  title={Cytoscape: a software environment for integrated models of biomolecular interaction networks},
  author={Shannon, Paul and Markiel, Andrew and Ozier, Owen and Baliga, Nitin S and Wang, Jonathan T and Ramage, Daniel and Amin, Nada and Schwikowski, Benno and Ideker, Trey},
  journal={Genome research},
  volume={13},
  number={11},
  pages={2498--2504},
  year={2003},
  publisher={Cold Spring Harbor Lab}
}
```

* Step 2: In your R Markdown document, add a code chunk that specifies the location of your .bib file and sets the bibliography style using the bibliography and biblio-style options, respectively.

```{}
---
title: "BCB420 - Computational Systems Biology"
subtitle: "Lecture 6 - Annotation Dataset and Intro to Pathway analysis"
author: "Ruth Isserlin"
date: "r Sys.Date()"
output:
  html_document:
    toc: true
    toc_depth: 2
bibliography: lecture6.bib
---
```

* Step 3: In the body of your R Markdown document, you can cite your references using the `@` symbol followed by the citation key, like this: `@cytoscape2003`. For example:

In his groundbreaking article, @cytoscape2003 proposed a theory of everything that has since revolutionized the field

# Reference 

